An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-09-21 report.tex
updates to report and forward-reduction diagram

    The right boundary ($f^{\prime}_{n-1}$) is treated by utilizing
    the negative complex-conjugate of the Fourier image of the stencil
    at $f^{\prime}_1$:

    \begin{equation}
        f^{\prime}_{n-1} + 2f^{\prime}_{n-2}
        =
        \frac{5f_{n-1} - 4f_{n-2} - f_{n-3}}{dx}
    \end{equation}

    The tridiagonal system that needs to be solved for the derivatives,
    then has an LHS of the form:

    \[ %\arraycolsep=4pt
     \begin{bmatrix}
         1&2\\
         1/4&1&1/4\\
         &1/4&1&1/4\\
         &&1/4&1&1/4\\
         &&&1/4&1&1/4\\
         &&&&&\ddots\\
         &&&&&&\ddots\\
         &&&&&&&\ddots\\
         &&&&&&&2&1
      \end{bmatrix}
    \]

    And an RHS of the form:

    \[
     \begin{bmatrix}
         -5f_1 + 4f_2 + f_3\\
         f_{3} - f_{1}\\
         f_{4} - f_{2}\\
         \vdots\\
         \vdots\\
         \vdots\\
         \vdots\\
         f_{n} - f_{n-2}\\
         5f_{n} - 4f_{n-1} - f_{3}
      \end{bmatrix}
    \]

@@

    GPUs afford massive parallelism by virtue of several (thousand)
    lightweight compute threads, referred to as \emph{work items}.
    The work items on a GPU are scheduled as \emph{blocks},
    which can be 1-D, 2-D, or 3-D.
    Each block is assigned to a \emph{streaming multiprocessor} (SM)
    on the GPU---the Tesla K20 GPU has 13 streaming multiprocessor---and
    all threads in a block are executed simultaneously by the SM.
    All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
    Access to global memory is efficient when concurrent threads
    access neighbouring memory locations
    (known as \emph{coalesced memory access}).
    All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
    Access to shared memory, in general, is much faster than access to global memory,
    but the amount of shared memory per SM is limited, typically a few dozen KB.
    Certain memory access patterns by concurrent threads to shared memory may result in
    serialization of the memory accesses, known as bank conflicts. These are detailed [here].

@@

    % left, lower, right, upper
    \begin{figure}[h]
    \begin{center}
    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
    \end{center}
    \caption{Problem domain}
    \label{fig:domain}
    \end{figure}

    \begin{figure}[ht]
    \begin{minipage}[b]{0.45\linewidth}
    \centering
    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
    \caption{Single block assigned to a process}
    \label{fig:block}
    \end{minipage}

    \hspace{0.5cm}
    \begin{minipage}[b]{0.45\linewidth}
    \centering
    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
    \caption{Single grid line within a block}
    \label{fig:grid-line}
    \end{minipage}
    \end{figure}

    The compact finite differences need to be evaluated
    at each grid point in the problem domain,
    shown in \ref{fig:domain}.
    Each grid point is represented as a cell in this representation.
    The domain has been shown as partitioned into blocks \ref{fig:one-block}.
    The blocks are three dimensional to facilitate
    the computation of derivatives in all three coordinate directions
    without incurring excessive communication costs.

    While evaluating the derivative in the x-direction (say),
    each block is assigned to a \emph{compute unit}, or a \emph{process}
    which may be a single core, a multi-core processor, or a GPU.
    Each block may be thought of as a collection of ``lines'' of grid points
    \ref{fig:one-line}---referred to as \emph{grid lines}---in the x-direction,
    and the entire domain may be thought of as a collection of such blocks.

    \subsection{Evaluation of the RHS}

        The RHS is evaluated by a 3 point stencil computation of
        the following form:

        \begin{equation*}
            r_i = \frac{3}{4dx} (f_{i+1} - f_{i-1})
        \end{equation*}

        This pointwise update can be performed easily in parallel
        for the internal points (points not close to the boundary of the process).
        Near the boundaries, the process will require information from its
        neighbours.

        Thus, each process will require information from its
        neighbouring processes every time the derivative needs to be evaluated.

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-domain.eps}
        \end{center}
        \caption{Distribution of a 2-D domain among processes}
        \label{fig:da-domain}
        \end{figure}

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-local-portion.eps}
        \end{center}
        \caption{Elements required by a process to evaluate RHS}
        \label{fig:da-local-portion}
        \end{figure}

        Thus, the RHS is evaluated in two steps:

        \begin{enumerate}
            \item Exchange boundary information with neighbouring processes
            \item Evaluate the RHS at each grid point (in parallel)
        \end{enumerate}

        The GPU is specialized for performing such a large number of independent
        computations as required in step 2.
        The RHS at each grid point is computing by a single work item
        (GPU thread).
        Therefore the number of work items scheduled is
        equal to the number of points in the local domain.

        Another popular strategy is to take advantage of
        shared memory, by launching 2-dimensional blocks of work-items,
        loading each into shared memory, and streaming the third dimension
        in registers.
        This approach is discussed in [ref], and is much more effective
        for larger stencils---as each thread in a block will access the same
        value several times, it is effective to use shared memory as explicitly
        controlled cache.

    \subsection{Solving the tridiagonal system}

        In the reference approach, a significantly larger portion of time is spent in solving
        the tridiagonal system in parallel.
        Because the problem does not exhibit the same degree of parallelism as
        the pointwise stencil update,
        the algorithms are necessarily more involved.

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/block-dimensions.eps}
        \end{center}
        \caption{Domain}
        \label{fig:block-dimensions}
        \end{figure}

        In the reference approach, the domain is decomposed into several blocks,
        and each block is assigned to a CPU core.
        Each core operates on a block with dimensions as shown
        in the figure \ref{fig:block-dimensions}.
        The computational pattern followed by each core for the steps listed in Section[SEC]
        is a triple nested loop over the grid points in the block,
        followed by a communication of the ``faces'' of the block with the other
        cores:

        \begin{listing}
        \begin{minted}{fortran}
           do j=1,n2
              do i=1,n1
                 x(1,i,j) = bet(1)*x(1,i,j)
                 do k = 2, n3
                    x(k,i,j) = (x(k,i,j) - a(k)*x(k-1,i,j))*bet(k)
                 end do
                 p_loc(i,j) = x(n3,i,j)
              end do
           end do

@@

        \end{minted}
        \caption{Computational pattern in reference approach}
        \end{listing}

        Conceptually, the innermost loop solves the tridiagonal system
        for a single ``line'' of grid points.

        In a multi-threaded environment,
        rather than having a single thread perform the local part
        of the computations for each line,
        it is beneficial to have several threads ``sweep'' through the domain,
        each performing the computations for a different tridiagonal system.
        This is shown in Figure \ref{fig:block-multi-threaded} for a 4-core processor.
        In each iteration of the inner-most loop,
        4 lines of grid points are solved.

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{50pt} {250pt} {50pt} {150pt}}, clip, height=300pt]{img/block-multi-threaded.eps}
        \end{center}
        \caption{Domain}
        \label{fig:block-multi-threaded}
        \end{figure}

        Algorithms like LU and the Thomas Algorithm are inherently sequential.
        While thread level parallelism can be used to solve several systems at once,
        the number of steps to solve each system remains large.

        Algorithms like Cyclic Reduction and Parallel Cyclic Reduction use
        several parallel processors to solve a tridiagonal system.
        Cyclic reduction

@@

    First, we discuss the cluster level parallelism.

@@

    \begin{equation} \label{eqn:reduced-system}
     \begin{bmatrix}
         x_{1,M}^{LH}   &   -1                                                          \\
         -1             &   x_{2,1}^{UH} &   x_{2,1}^{LH}                               \\
                        &   x_{2,M}^{LH} &   x_{2,M}^{LH}  &  -1                        \\
         &              &   -1           &   x_{3,M}^{UH}  &  x_{1,M}^{LH}              \\
         &              &   &                x_{3,M}^{UH}  &  x_{1,M}^{LH} &    -1      \\
         &              &   &            &   &             \ddots   \\
         &              &   &            &   &             &        \ddots \\
         &  &   &   &   &   &   & -1 x_{P,1}^{UH}
      \end{bmatrix}
    \begin{bmatrix}
        \xi_1^{LH} \\
        \xi_2^{UH} \\
        \xi_2^{LH} \\
        \xi_3^{UH} \\
        \xi_3^{LH} \\
        \vdots \\
        \vdots \\
        \xi_P^{UH} \\
     \end{bmatrix}
     =
     \begin{bmatrix}
         x_{1,M}^R \\
         x_{2,1}^R \\
         x_{2,M}^R \\
         x_{3,1}^R \\
         x_{3,M}^R \\
         \vdots \\
         \vdots \\
         x_P^R \\
      \end{bmatrix}
  \end{equation}

@@

    \begin{enumerate}
        \item Each process solves the three equations
        \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}.
        \item Each process communicates the parameters required to
        assemble the ``reduced'' system eqn:reduced-system
        \item The reduced system is solved for the coupling coefficients
        $\xi_i^{UH}$ and $_i^{LH}$, which are sent to their respective processes
        (if required)
        \item Each process solves \ref{eqn:linear-combo} for the local part
        of the solution to the global system.
    \end{enumerate}

@@

    \begin{enumerate}
        \item Each process is involved in solving \emph{several} equations,
            one for each grid line.
        \item
        \item
    \end{enumerate}

    * Each process is involved in solving a number of equations, and not a single one
    * The left hand side for all the equations is the same,
        - LHS of \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}
        - RHS of \ref{eqn:local-eqn-2} and \ref{eqn:local-eqn-3} are the same
        - The above two equations need only be solved for a single grid line
    * Only equation \ref{eqn:local-eqn-3} is solved for every grid line,
    which amounts to solving the same equation for several different RHS

    Cyclic Reduction
    Precomputing coefficients in cyclic reduction
    OpenCL implementations
    Shared memory and associated issues
    Repeated kernel invocations
    Final kernel listings

    Overall solution strategy

\section{Code design}

\section{Results and performance comparison}
    Notes on performance comparison
        - Fair comparisons
        - Memory transfers
        - Time spent on writing code
        - Compiler optimizations
    Problem size v/s solution time
    Breakup of time taken for various problem sizes and nprocs
    Results for CPU

    \subsection{Results for a single node}
    The results for a single node bring up some issues in performance evaluation.
    First, it is clear that the major cost for the GPU is not in the computational
    kernels, but rather in transferring data between host (CPU) and device (GPU) buffers.
    In the context of a larger piece of software,
    if only the computation of the compact finite difference was offloaded
    to the GPU, then considering this cost is absolutely essential.
    However, if the rest of the problem is also written to be solved on the GPU,
    then there is no need for transfers to and from the host at each
    time step.

    \subsection{Results for multi-node}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-x.eps}
\end{center}
\caption{Parallel x}
\label{fig:parallel-x}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-yz.eps}
\end{center}
\caption{Parallel yz}
\label{fig:parallel-yz}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/parallel-both.eps}
\end{center}
\caption{Parallel-both}
\label{fig:parallel-both}
\end{figure}

>>>>

    The right boundary ($f^{\prime}_{n}$) is treated by utilizing
    the negative complex-conjugate of the Fourier image of the stencil
    at $f^{\prime}_1$:

    \begin{equation}
        f^{\prime}_{n} + 2f^{\prime}_{n-1s}
        =
        \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
    \end{equation}

    The tridiagonal system that needs to be solved for the derivatives,
    then has the form:

    \begin{equation} \label{eqn:compact-tridiagonal-system}
     \begin{bmatrix}
         1&2\\
         1/4&1&1/4\\
         &1/4&1&1/4\\
         &&1/4&1&1/4\\
         &&&1/4&1&1/4\\
         &&&&&\ddots\\
         &&&&&&\ddots\\
         &&&&&&&\ddots\\
         &&&&&&&2&1
      \end{bmatrix}
      \begin{bmatrix}
          f^{\prime}_1 \\
          f^{\prime}_2 \\
          f^{\prime}_3 \\
          \vdots \\
          \vdots \\
          \vdots \\
          \vdots \\
          f^{\prime}_{n-1} \\
          f^{\prime}_n
       \end{bmatrix}
     =
     \begin{bmatrix}
         -5f_1 + 4f_2 + f_3\\
         f_{3} - f_{1}\\
         f_{4} - f_{2}\\
         \vdots\\
         \vdots\\
         \vdots\\
         \vdots\\
         f_{n} - f_{n-2}\\
         5f_{n} - 4f_{n-1} - f_{n-2}
      \end{bmatrix}
    \end{equation}

@@

    GPUs afford massive parallelism by virtue of several (thousand)
    lightweight compute \emph{threads} or \emph{work-items}.
    The threads on a GPU are scheduled as \emph{blocks},
    which can be 1-D, 2-D, or 3-D.
    Each block is assigned to a \emph{streaming multiprocessor} (SM)
    on the GPU---the Tesla K20 GPU has 13 streaming multiprocessor---and
    all threads in a block are executed simultaneously by the SM.
    All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
    Access to global memory is efficient when concurrent threads
    access neighbouring memory locations
    (known as \emph{coalesced memory access}).
    All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
    Access to shared memory, in general, is much faster than access to global memory,
    but the amount of shared memory per SM is limited, typically a few dozen KB.
    Certain memory access patterns by concurrent threads to shared memory may result in
    serialization of the memory accesses, known as bank conflicts. These are detailed [here].

@@

    % left, lower, right, upper
    \begin{figure}[h]
    \begin{center}
    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/domain.eps}
    \end{center}
    \caption{Problem domain}
    \label{fig:domain}
    \end{figure}
    %
    \begin{figure}[ht]
    \begin{minipage}[b]{0.40\linewidth}
    \begin{center}
    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
    \end{center}
    \caption{A subdomain}
    \label{fig:block}
    \end{minipage}
    %
    \hspace{0.5cm}
    \begin{minipage}[b]{0.40\linewidth}
    \begin{center}
    \includegraphics[trim={{50pt} {150pt} {100pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
    \end{center}
    \caption{Single grid line within a subdomain}
    \label{fig:grid-line}
    \end{minipage}
    \end{figure}
    %
    \begin{figure}[h]
    \begin{center}
    \includegraphics[trim={{100pt} {250pt} {100pt} {150pt}}, clip, height=200pt]{img/block-dimensions.eps}
    \end{center}
    \caption{Domain}
    \label{fig:block-dimensions}
    \end{figure}

    \pagebreak

    The compact finite differences need to be evaluated
    at each grid point in the problem domain,
    shown in \ref{fig:domain}.
    Each grid point is represented as a cell in this representation.
    The domain has been shown as partitioned into subdomains \ref{fig:block}.
    The subdomains are three dimensional to facilitate
    the computation of derivatives in all three coordinate directions
    without incurring excessive communication costs.

    While evaluating the derivative in the x-direction (say),
    each subdomain is assigned to a \emph{compute unit}, or a \emph{process}
    which may be a single core, a multi-core processor, or a GPU.
    Each subdomain may be thought of as a collection of ``lines'' of grid points
    \ref{fig:grid-line}---referred to as \emph{grid lines}---in the x-direction,
    and the entire domain may be thought of as a collection of such subdomains.

    In the reference approach, each subdomain is assigned to a CPU core.
    Each core operates on a subdomain with dimensions as shown
    in the figure \ref{fig:block-dimensions}.
    The computational pattern followed by each core for the steps listed in Section[SEC]
    is a triple nested loop over the grid points in the subdomain,
    followed by a communication of the ``faces'' of the subdomain with the other
    cores:

    \begin{listing}
    \begin{minted}{fortran}

       do j=1,n2
          do i=1,n1
             x(1,i,j) = bet(1)*x(1,i,j)
             do k = 2, n3
                x(k,i,j) = (x(k,i,j) - a(k)*x(k-1,i,j))*bet(k)
             end do
             p_loc(i,j) = x(n3,i,j)
          end do
       end do

@@

    \end{minted}
    \caption{Computational pattern in reference approach}
    \label{listing:triple-loop}
    \end{listing}

    Conceptually, the innermost loop solves the tridiagonal system
    for a single grid line,
    while the outer loops navigate the different grid lines.
    The evalution of such triple loops are found to be the most time-consuming steps
    in the solution process,
    and it is desirable to introduce a second level of parallelism to replace these loops.
    Broadly speaking, there are two forms of parallelism that can be exploited:

    \textbf{Thread-level parallelism} \\
        In a multi-threaded environment,
        rather than having a single thread perform the local part
        of the computations for each line,
        it is beneficial to have several threads ``sweep'' through the subdomain,
        each performing the computations for a different tridiagonal system.
        This is a sort of \emph{thread parallel} operations (Figure \ref{fig:thread-parallel}).
        Algorithms that are inherently sequential,
        such as the Thomas algorithm lend themselves easily to thread-level parallelism
        With reference to the triple-loop in Listing 1,
        algorithms employing thread-level parallelism reduce the
        size of the outer two loops.

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/thread-parallel.eps}
        \end{center}
        \caption{Parallel x}
        \label{fig:thread-parallel}
        \end{figure}

    \textbf{Data parallelism} \\
        Algorithms like Cyclic Reduction and Parallel Cyclic Reduction
        use parallel processors to perform
        perform the same operation at different points in the tridiagonal system.
        They are said to exhibit \emph{data parallelism} (Figure \ref{fig:data-parallel}).
        Although these algorithms perform a larger number of computations
        per step compared to the Thomas algorithm,
        they can take far fewer steps to run,
        depending on the number of available parallel processors.
        With reference to the triple-loop in Listing 1,
        algorithms employing data parallelism reduce the size of the innermost loop.

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/data-parallel.eps}
        \end{center}
        \caption{Parallel yz}
        \label{fig:data-parallel}
        \end{figure}

    In this work, we evaluate the performance of a variant of Cyclic Reduction
    on the GPU that is both data-parallel and thread-parallel,
    and a thread-parallel implementation of the Thomas Algorithm on
    multi-core CPUs.

    Both implementations are compared against the reference implementation
    described in Section [SEC].
    The thread-parallel implementation of Thomas Algorithm is
    straightforward.
    However, implementation Cyclic Reduction on GPUs involves several issues,
    and our algorithm is described in detail in Section [SEC].
    %
    % \subsection{Evaluation of the RHS}
    %
    %     The RHS is evaluated by a 3 point stencil computation of
    %     the following form:
    %
    %     \begin{equation*}
    %         r_i = \frac{3}{4dx} (f_{i+1} - f_{i-1})
    %     \end{equation*}
    %
    %     This pointwise update can be performed easily in parallel
    %     for the internal points (points not close to the boundary of the process).
    %     Near the boundaries, the process will require information from its
    %     neighbours.
    %
    %     Thus, each process will require information from its
    %     neighbouring processes every time the derivative needs to be evaluated.
    %
    %     \begin{figure}[h]
    %     \begin{center}
    %     \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-domain.eps}
    %     \end{center}
    %     \caption{Distribution of a 2-D domain among processes}
    %     \label{fig:da-domain}
    %     \end{figure}
    %
    %     \begin{figure}[h]
    %     \begin{center}
    %     \includegraphics[trim={{100pt} {250pt} {100pt} {200pt}}, clip, height=200pt]{img/da-local-portion.eps}
    %     \end{center}
    %     \caption{Elements required by a process to evaluate RHS}
    %     \label{fig:da-local-portion}
    %     \end{figure}
    %
    %     Thus, the RHS is evaluated in two steps:
    %
    %     \begin{enumerate}
    %         \item Exchange boundary information with neighbouring processes
    %         \item Evaluate the RHS at each grid point (in parallel)
    %     \end{enumerate}
    %
    %     The GPU is specialized for performing such a large number of independent
    %     computations as required in step 2.
    %     The RHS at each grid point is computing by a single work item
    %     (GPU thread).
    %     Therefore the number of work items scheduled is
    %     equal to the number of points in the local domain.
    %
    %     Another popular strategy is to take advantage of
    %     shared memory, by launching 2-dimensional blocks of work-items,
    %     loading each into shared memory, and streaming the third dimension
    %     in registers.
    %     This approach is discussed in [ref], and is much more effective
    %     for larger stencils---as each thread in a block will access the same
    %     value several times, it is effective to use shared memory as explicitly
    %     controlled cache.

@@

    The size of the problems solved using the high-order compact finite
    difference schemes discussed is usually much larger than can be accomodated
    on a single compute node.
    The problem domain is thus divided into ``subdomains'' and distributed among
    several processes.
    First, we discuss this cluster level parallelism.

@@

    \begin{equation} \label{eqn:reduced-system}
     \begin{bmatrix}
         x_{1,M}^{LH}   &   -1                                                          \\
         -1             &   x_{2,1}^{UH} &   x_{2,1}^{LH}                               \\
                        &   x_{2,M}^{LH} &   x_{2,M}^{LH}  &  -1                        \\
         &              &   -1           &   x_{3,M}^{UH}  &  x_{1,M}^{LH}              \\
         &              &   &                x_{3,M}^{UH}  &  x_{1,M}^{LH} &    -1      \\
         &              &   &            &   &             \ddots   \\
         &              &   &            &   &             &        \ddots \\
         &  &   &   &   &   &   & -1 x_{P,1}^{UH}
      \end{bmatrix}
    \begin{bmatrix}
        \xi_1^{LH} \\
        \xi_2^{UH} \\
        \xi_2^{LH} \\
        \xi_3^{UH} \\
        \xi_3^{LH} \\
        \vdots \\
        \vdots \\
        \xi_P^{UH} \\
     \end{bmatrix}
     = -
     \begin{bmatrix}
         x_{1,M}^R \\
         x_{2,1}^R \\
         x_{2,M}^R \\
         x_{3,1}^R \\
         x_{3,M}^R \\
         \vdots \\
         \vdots \\
         x_P^R \\
      \end{bmatrix}
  \end{equation}

@@

    \begin{enumerate}
        \item Each process solves the three equations
        \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}.
        \item Each process communicates the parameters required to
        assemble the ``reduced'' system eqn:reduced-system
        \item The reduced system is solved for the coupling coefficients
        $\xi_i^{UH}$ and $xi_i^{LH}$, which are sent to their respective processes
        (if required)
        \item Each process solves \ref{eqn:linear-combo} for the local part
        of the solution to the global system.
    \end{enumerate}

@@

    \begin{enumerate}
        \item Each process solves equations \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}
            for \emph{each grid line} in its subdomain.
        \item The left-hand side of these equations is the same
            for each grid line, coming from equation \ref{eqn:compact-tridiagonal-system}).
        \item The right-hand side for equations \ref{eqn:local-eqn-1} and
            \ref{eqn:local-eqn-2} are the same for each grid line.
            Thus, the equations need only be solved for a single grid line.
        \item Only \ref{eqn:local-eqn-3} is evaluated for each grid line
            in the subdomain. This is expected to be the most computationally expensive step.
    \end{enumerate}

    \subsection{Cyclic reduction}

        The classic Cyclic Reduction algorithm consists of two phases:
        forward reduction and backward substitution.
        The forward substitution phase begins with a system of size $N$,
        and by expressing every even-indexed equation $i$ as a linear
        combination of equations $i$, $i-1$ and $i+1$, reduces it to a
        system of size $N/2$.
        The process is repeated until a system of 2 equations in 2 unknowns
        is left.

        For each even-indexed $i$, we define:

        \begin{equation*}
        k_1 = \frac{a_i}{b_{i-1}},
        k_2 = \frac{c_i}{b_{i+1}}
        \end{equation*}

        Then, we update the values of $a$, $b$, $c$ and $d$ as:

        \begin{align} \label{eqn:forward-reduction}
        & a^{\prime}_i = -a_{i-1}k_1 & \\
        & b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
        & c^{\prime}_i = -c_{i+1}k_2 & \\
        & d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 &
        \end{align}

        The two equations are solved, yielding $x_N$ and $x_{N/2}$.
        In the backward substitution phase, every odd-indexed unknown $x_i$
        is solved for by substituting the known values of $x_{i-1}$ and $x_{i+1}$:

        \begin{align} \label{eqn:back-substitution}
        x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
        \end{align}

        At each step, for the last index $i$, the forward reduction step is
        instead:
        \begin{align} \label{eqn:forward-reduction-right}
            & a^{\prime}_i = -a_{i-1}k_1 & \\
            & b^{\prime}_i = b_i - c_{i-1}k_1 & \\
            & d^{\prime}_i = d_i - d_{i-1}k_1 &
        \end{align}

        And the backward reduction step for $i=1$ is instead:
        \begin{align} \label{eqn:back-substitution-right}
        x_i = \frac{d^{\prime}_i - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
        \end{align}

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{50pt} {300pt} {100pt} {50pt}}, clip, height=200pt]{img/forward-reduction.eps}
        \end{center}
        \caption{Cyclic reduction}
        \label{fig:forward-reduction}
        \end{figure}

        The cyclic reduction algorithm is well suited to the massively
        parallel architecture of the GPU.
        In the most straightforward implementation,
        we may launch a number of threads equal to
        half the number of equations,
        and assign each thread the task of
        solving the equations \ref{eqn:forward-reduction} for each even-indexed
        $i$.
        We can then perform the successive steps of forward reduction
        by halving the number of threads at each step.
        The backward substitution phase is completed in a similar way,
        with the number of threads doubling at each step.

        This naive approach leads to the following issues:

        \begin{enumerate}
            \item Most of the threads are inactive,
            except at the very beginning of forward reduction
            and the very end of backward substitution.

            \item The later steps of forward reduction and the first few
            steps of backward substitution involve memory accesses with
            large strides (see Figure \ref{fig:forward-reduction})
            If the threads are accessing global memory,
            this can lead to poor performance.
        \end{enumerate}

        The first issue is fairly straightforward to tackle.
        The OpenCL API allows \emph{scheduling} of a large number of
        threads---far more than the number of threads actually available
        on the computing device.
        Rather than scheduling threads for a single system (grid line),
        we may schedule threads to perform
        forward reduction steps for every grid line.

        To avoid non-coalesced accesses to global memory,
        we can instead have threads working with the fast, on-chip,
        \emph{shared memory}. In this case, the threads in each
        \emph{block} read data from global memory on to the block's
        shared memory.
        The threads then perform cyclic reduction on shared memory
        and write the data back to global memory.
        Conceptually, this approach has the advantage of avoiding
        non-coalesced global memory accesses, and facilitates data reuse.
        However, as described in [FastTridiagonalSolvers], the method
        suffers from so-called \emph{bank conflicts} in which,
        for certain memory access patters,
        data access by concurrent threads from shared memory may be serialized.
        This can seriously affect performance.
        Several approaches for avoiding bank conflicts have been proposed,
        this is at the cost of additional code complexity.
        Besides, an approach using shared memory suffers from the penalty
        of \emph{synchronization} between blocks and inactive threads.

    \subsection{Precomputing scheme}

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{50pt} {300pt} {100pt} {50pt}}, clip, height=200pt]{img/forward-reduction-coefficients.eps}
        \end{center}
        \caption{Forward reduction: indices with identical coefficients are shown in color}
        \label{fig:forward-reduction-coefficients}
        \end{figure}

        We introduce a variant of cyclic reduction,
        specialized for constant tridiagonal systems, that:

        \begin{enumerate}
        \item Reduces the number of accesses to global memory
        \item Reduces the memory requirement
        \item Reduces the number of computations in the forward reduction step
        \end{enumerate}

        We recall the form of the tridiagonal matrix:

        \[
        \begin{bmatrix}
            1&2\\
            1/4&1&1/4\\
            &1/4&1&1/4\\
            &&1/4&1&1/4\\
            &&&1/4&1&1/4\\
            &&&&&\ddots\\
            &&&&&&\ddots\\
            &&&&&&&\ddots\\
            &&&&&&&2&1
         \end{bmatrix}
         \]

        The matrix is the same for each grid line in the subdomain.
        Thus,
        the first three equations in \ref{eqn:forward-reduction}
        are the same for every grid line.
        For each forward reduction step,
        the parameters $a^{\prime}_i$, $b^{\prime}_i$ and $c^{\prime}_i$
        can be pre-computed.
        For a general tridiagonal system, this would require storage for
        $N/2 + N/4 + ... = 2N-2$ coefficients.
        However, it is readily apparent that,
        with the exception of the boundaries (near the first and last equations),
        the coefficients $k_1$, $k_2$,
        $a_i$, $b_i$ and $c_i$ are the same for each $i$.

        The size of the precomputed coefficient arrays is then $log_2(N)-1$
        (with an additional element for $a_N$ and $b_N$ as described below).
        For event modestly sized $N$ such as 512, this represents large savings
        in the memory requirements for the coefficient arrays.

        More importantly, at each forward reduction step,
        the number of non-coalesced memory accesses is greatly reduced.
        With the above precomputed coefficients,
        the forward reduction step now involves solving only a single equation
        for $d_i$ at each $i$.

        \begin{align} \label{eqn:forward-reduction-right}
            & a^{\prime}_i = -a_{i-1}k_1 & \\
            & b^{\prime}_i = b_i - c_{i-1}k_1 & \\
            & d^{\prime}_i = d_i - d_{i-1}k_1 &
        \end{align}

        The backward reduction step benefits from improved access to
        global memory.

        \subsubsection*{Boundary conditions}

        The tridiagonal systems for all grid lines are the same for all subdomains,
        except near the boundaries.
        From \ref{eqn:forward-reduction}, it is clear that
        to handle the boundaries, three additional sets of coefficients
        need to be stored:

        \begin{enumerate}
         \item The values for $b_1$ at each step
         \item The values for $k_1$ evaluated at $i=1$ for each step
         \item The values for $k_1$ evaluated at $i=n$ for each step,
            where $n$ is the last index of the reduction step.
        \end{enumerate}

        The values of $a_N$ and $b_N$ are required only for the
        final step (2-by-2 solve) of forward reduction.
        These two scalar values may be stored separately.

\section{Results and performance comparison}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=300pt]{img/data-and-thread-parallel.eps}
\end{center}
\caption{Parallel-both}
\label{fig:parallel-both}
\end{figure}
